\chapter{Background and Related Work} \label{chap:background-related-work}
This chapter reviews prior work related to \textit{IEC 61499}-based development and the used \acp{IDE}. The goal is to better understand the landscape, which will then allow the critical review of the state of the art. By providing a comprehensive analysis of the current landscape of development environments for \acp{CPPS} and the \textit{IEC 61499} standard, we are able to categorise and critically examine the existing solutions, while looking for possible gaps in the ecosystem.

\section{Methodology}
A targeted literature review was carried out, focusing on studies addressing \textit{IEC 61499} development tools, function blocks, \apc{CPPS} and development environments, to then critically reflect upon their findings to help the writing of this dissertation and the planning of the final product.

\subsection{Search and Selection}
The search was conducted in digital libraries \textit{ScienceDirect}, \textit{IEEE Xplore}, \textit{ACM Digital Library}, among others, as well as research paper aggregators like \textit{Semantic Scholar} and \textit{Scopus} using combinations of the following terms:

\begin{itemize}
  \item \textit{IEC 61499}
  \item \ac{FBD}
  \item \ac{IDE}
  \item \ac{CPPS}
\end{itemize}

Following the search, a manual review process was conducted to refine and validate the results. Titles and abstracts were first screened to understand the relevance of the articles with respect to this dissertation. The publications were then examined in greater detail through full-text review, the reference lists of selected papers were manually inspected to identify relevant works that may not have been captured by the initial keyword search.

Articles mentioning new development environment proposals \cite{sorokin_novel_2022} were immediately included as their novel ideas were of great interest for a gap analysis on this ecosystem. These papers usually already contained a gap analysis of their own, which allowed for a literature review that was consistent with the prevalent opinion of the scientific community.

Furthermore, terms related to the state of the art that were discovered mid search were also included in some queries. The objectives of doing this were to get a better personal understanding of the scene, as well as to find interesting references that could be included in this dissertation to backup the written claims. Some of the terms used were:

\begin{itemize}
  \item 4diac
  \item remote
  \item collaboration
  \item real-time
\end{itemize}

\section{Distributed Function Block Engineering}
Modern manufacturing and industrial automation are increasingly characterised by the transition from monolithic centralised control architectures towards distributed modular systems. The industry is moving towards a paradigm in which individual parts possess intelligence and communicate with one another. This shift is driven by the demands of Industry 4.0, which calls for flexibility and interoperability as key properties of the production infrastructure \cite{folgado_review_2024}. \citeauthor{monostori_cyber-physical_2014} \cite{monostori_cyber-physical_2014} provides a comprehensive analysis of \ac{CPPS}, examining the theoretical foundations and practical implications of tightly coupling computational intelligence with physical manufacturing processes. The author explores how cyber-physical integration enables real-time monitoring, adaptive control and autonomous decision making in production systems, framing \ac{CPPS} as a cornerstone of the fourth industrial revolution. This work establishes the concepts, wording and architectural principles that help understand much of the research in this area and it serves as a reference for understanding why the flexible distributed control is necessary to realize the vision of \ac{CPPS}.

The \textit{IEC 61499} standard was developed precisely to address these requirements at the software level. This architecture defines a domain-specific modelling language for developing distributed \ac{ICS}. It does so by extending and superseding the older \textit{IEC 61131-3} standard, which was designed for centralised \ac{PLC}-based control, providing next-generation systems with an architecture that enables distributed control \cite{bencherki_iec_2024}. The architecture encourages flexible reuse of components as well as reliable data exchange.

In \ac{FBD}, function blocks are interconnected into networks that form an application, which can then be distributed across multiple devices \cite{liang_kind_2008}, a property that makes \ac{FBD} a natural fit for \ac{CPPS} deployments where hardware must collaborate to execute an unified control program. \citeauthor{torres_literature_2025} \cite{torres_literature_2025} conducted a literature review examining how \textit{IEC 61499} can be leveraged specifically for \ac{CPPS} development, mentioning possible enhancements with the use of a cloud-based \ac{SOA} approach. Their paper examines the current state of adoption of the standard within industrial automation research, identifying recurring architectural patterns in \ac{CPPS} implementations.

Conversely, \citeauthor{pinto_digital_2023} \cite{pinto_digital_2023} presented an industrial case study showing that function block oriented approaches improve system modularity and reduce development effort compared to traditional procedural programming. Their study documents the migration of a production cell from a sequential control architecture to an \textit{IEC 61499} function block model using the \textit{DINASORE} \ac{RTE}, using integration time as the main validation metric and demonstrating improved reconfigurability when production requirements changed. The results show that the encapsulation of behaviour and communication interfaces within function blocks leads to more maintainable and extensible software.

% TODO talk about fbd

\section{IEC 61499 Development Environments}
Currently, the development ecosystem for \textit{IEC 61499} is composed almost exclusively of desktop-based \acp{IDE}. These tools are typically monolithic applications installed locally on engineering workstations, requiring each member of a development team to independently install, configure and maintain the software on their own machine. The tools themselves tend to be resource intensive, which creates demands for the hardware, making them less accessible to users working on modest, shared or virtual machines. There is also the issue of platform dependency, as a niche and focused field, some \acp{IDE} are not crossplatform, forcing users to use virtual machines to be able to use them, further intensifying performance issues.

\subsection{Open-Source}
The most prominent open source implementation is the \textit{4diac IDE} (based on the \textit{Eclipse} framework) which supports a variety of \acp{RTE} such as \textit{4diac FORTE} and \textit{DINASORE}. This \ac{IDE} is very feature rich at the cost of increased complexity, resulting in a steep learning curve, which makes it difficult to use \cite{weber_usability_2019}. The \textit{Eclipse} foundation on which it is built is itself a heavyweight platform and while this grants access to a broad plugin ecosystem, it also means that the \ac{IDE} inherits the framework's architectural complexity. Navigating the interface to accomplish basic tasks such as creating a new function block type, designing the flow and deploying to a \ac{RTE} can require substantial initial investment before productive work can begin.

\citeauthor{zoitl_developing_2013} \cite{zoitl_developing_2013} demonstrated the use of \textit{4diac IDE} to develop modular and reusable \textit{IEC 61499} applications, highlighting the benefits of flexibility and adaptability in industrial settings. The paper provides a practical perspective on the strengths and limitations of this \ac{IDE}.

An alternative is \textit{FBME}, which is a much simpler modelling environment based on the \textit{JetBrains MPS} language workbench. This \ac{IDE} implements \textit{IEC 61499} concepts through a set of integrated \acp{DSL}. The main contribution of this work is the research on applying model-driven development principles to the architecture of the \ac{IDE} itself \cite{sorokin_novel_2022}, resulting in a very modular and extensible product.

\subsection{Commercial}
The commercial side is occupied by \textit{EcoStruxure Automation Expert}, \textit{ISaGRAF Workbench} \cite{vyatkin_comparisons_2008}, tools that have their own built-in proprietary \ac{RTE}. This tight coupling between the engineering tool and the runtime gives commercial vendors significant control over the end development experience, which typically results in polished, production-ready tooling with strong vendor support. However, it also introduces considerable lock-in: projects developed in one commercial environment are not easily portable to a different runtime or \ac{IDE} and the proprietary nature of the underlying \ac{RTE} limits the ability to inspect, extend, or adapt behaviour.

Commercial tools are also constrained by their licensing models. Per seat licensing is common, meaning that the cost of the toolchain scales directly with the size of the team and can not be viable for smaller organisations, academic research groups or early stage projects. This creates a tiered ecosystem in which well resourced industrial organisations have access to mature tooling, while smaller teams are restricted to the open source alternatives, each of which carries its own set of limitations as described above.

\subsection{Local-first Development} \label{sec:localfirst}
All of the tools mentioned previously share a fundamental design decision: they play around a local-first development model, in which the project state lives on a single workstation. While this model has historically been the norm in industrial automation tooling, it introduces a set of limitations that become problematic as \ac{CPPS} projects grow in scale and involve more distributed engineering teams.

The most immediate consequence is the absence of real-time collaboration. When two engineers need to work on the same \textit{IEC 61499} application simultaneously, they must coordinate manually, typically by dividing the project into separate files or modules and merging their changes afterwards. These tools also require the \ac{IDE} to maintain awareness of the deployment target's state. Without a shared backend, each engineer must independently configure their local environment to point to the correct runtime endpoints and therefore there is no single source of truth for the current deployment state of the system.

\section{Collaborative Modelling}
As projects grow in scale and involve distributed engineering teams, the question of how to coordinate concurrent modifications to a shared application model becomes central \cite{seidl_collaborative_2018}. Version control through \textit{Git} is inadequate in this domain as it operates on a line based difference algorithm designed for plain text source code. In the general case, it is not safe to merge structured model files using plain line based merge operations, as they can produce structurally consistent but semantically incorrect models . \textit{IEC 61499} applications are stored as \textit{XML} documents whose semantic integrity depends on relationships between elements. A valid merge operation can still result in a broken application graph.

Approaches to real-time concurrent editing, \ac{OT} and \ac{CRDT} offer a fundamentally different model of collaboration from the workflow of version control systems. Rather than requiring engineers to work in isolation and later reconcile divergent states, these techniques allow all participants to apply changes optimistically to their local replica, with the system guaranteeing that every replica converges to the same state. This shift ensures that inconsistencies are resolved continuously and automatically, rather than discovered only at integration time, enabling a concurrent engineering experience over shared structured artefacts such as \textit{IEC 61499} application graphs.

\section{Evolution of Web-Based Tools}
The software engineering domain has seen a massive shift from desktop \acp{IDE} to cloud based environments. This trend is now influencing the domain of \ac{ICS} and while recent studies have proposed a \ac{SOA} to integrate \textit{IEC 61499} with cloud layers \cite{torres_literature_2025}, the engineering tools themselves remain still desktop bound.

In general software development, tools like \textit{Visual Studio Code} have started offering web based counterparts to their desktop applications \cite{noauthor_vscode_nodate} in search of a more flexible user experience. Running entirely within the browser, these environments allow engineers to open, edit and navigate codebases from any device without installation. They integrate naturally with cloud hosted codebases and \ac{CI} pipelines. In fact, newer development tools like \textit{Zed} make it a key goal of their products to have web platform support \cite{noauthor_zed_nodate}, treating browser delivery not as an afterthought but as a primary design constraint. % TODO dont like zed as an example

Web based editors are also often easier to extend with real-time collaboration because they operate within a network architecture, leverage standardised browser communication technologies such as WebSockets and naturally integrate with centralised synchronisation services. These characteristics reduce many of the distribution and interoperability challenges that desktop applications must address separately \cite{schmid_real-time_2014}.

The move towards web based tooling is not limited to software development environments. Across a wide range of professional domains, browser-native applications have gradually replaced traditional desktop software. Collaborative document editors such as \textit{Google Docs} allow multiple users to edit the same document simultaneously, with changes synchronised in real time. Similarly, design and prototyping platforms such as \textit{Figma} have demonstrated that even complex graphical workflows can be delivered entirely through the browser while supporting collaboration and centralised asset management. Drawing tools like \textit{Excalidraw} and \textit{draw.io} further illustrate how browser based applications can provide sophisticated modelling and visualisation capabilities without requiring local installation, while providing live collaboration features that their desktop predecessors did not have.
% TODO mention they use json underneath

A common characteristic of these platforms is that they treat collaboration as an architectural design rather than an addon. By maintaining a shared representation of the artefact on a central service, users can concurrently make changes with minimal configuration. This contrasts with many traditional desktop applications, where collaboration often relies on exchanging files with separate version control and synchronisation mechanisms. As a result, web based tools have created a new expectation on collaboration.

Some studies explore remote web based execution and editing of \ac{FBD}. \citeauthor{rohat_remote_2014} \cite{rohat_remote_2014} proposed an early approach to remote monitoring and interaction with \textit{IEC 61499} applications through a web interface, identifying the practical benefits of browser access in distributed industrial settings. While the work predates many modern web technologies, it established the motivation for moving development and monitoring capabilities to the browser. More recently, \citeauthor{lyu_towards_2024} \cite{lyu_towards_2024} revisited this direction by proposing fully remote cloud based platform for \textit{IEC 61499} development, embedding \textit{EcoStruxure AE} in a frontend web interface. This way they were able to move all the processing into a centralised cloud platform.

More generally, \citeauthor{sorokin_towards_2023} \cite{sorokin_towards_2023} proposes a new architecture for the development of web-based \ac{DSL} tools, grazing on the field of of \textit{IEC 61499} engineering by using \textit{FBME} as an example of how it can be migrated into a distributed collaborative web environment. While this paper does not provide the groundwork and validation necessary for the development of specifically a \textit{IEC 61499} tool, it does provide good general insights of how its architecture could be designed.

\section{Comparative Analysis and Research Gap}
% TODO intro



\newcommand{\cyes}{\textcolor{teal}{\checkmark}}
\newcommand{\cno}{\textcolor{purple}{$\times$}}
\begin{table}[htbp]
\centering
\caption{Comparison of IEC 61499 development environments and proposals.}
\label{tab:ide-comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llp{2.5cm}ccp{5cm}@{}}
\toprule
\textbf{Tool} & \textbf{Type} & \textbf{Platform} & \textbf{\shortstack{Remote\\deployment}} & \textbf{\shortstack{Real-time\\collaboration}} & \textbf{Comments} \\
\midrule
4diac IDE \cite{zoitl_developing_2013}
  & Open source & Desktop
  & \cno & \cno
  & Steep learning curve \\[6pt]
FBME \cite{sorokin_novel_2022}
  & Open source & Desktop
  & \cno & \cno
  & Very modular \\[6pt]
EcoStruxure AE
  & Commercial & Desktop
  & \cno & \cno
  & Proprietary ecosystem \\[6pt]
ISaGRAF Workbench \cite{vyatkin_comparisons_2008}
  & Commercial & Desktop
  & \cno & \cno
  & Licensing model \\[6pt]
\citeauthor{rohat_remote_2014}\ \cite{rohat_remote_2014}
  & Proposal & Web (monitoring only)
  & \cyes & \cno
  & No remote modelling capabilities \\[6pt]
\citeauthor{lyu_towards_2024}\ \cite{lyu_towards_2024}
  & Proposal & Web (embedded desktop IDE)
  & \cyes & \cno
  & Does not take advantage of web technologies for the modelling tool \\
\citeauthor{sorokin_towards_2023}\ \cite{sorokin_towards_2023}
  & Proposal & Web
  & \cyes & \cyes
  & Designs a general architecture for creating web \ac{DSL} tooling  \\
PTERO\
  & New Proposal & Web
  & \cyes & \cyes
  & - \\
\bottomrule
\end{tabular}%
}
\end{table}

Current \textit{IEC 61499} tools face significant limitations that hinder modern \ac{CPPS} development. These gaps paint a picture of an ecosystem that has matured in terms of standards compliance, but has lagged considerably in adapting to the workflows and expectations of contemporary engineering teams. The following points describe the most significant gaps:

% TODO What other gaps?
% TODO add references
\begin{itemize}
    \item \textbf{Platform Dependency:} Existing solutions are monolithic desktop applications, lacking the accessibility and portability that we have come to expect. Engineers are required to install and maintain native software on every machine from which they intend to work, creating friction at the start of projects and compounding over time as the team grows or its members change.

    \item \textbf{Absence of Real-Time Collaboration:} None of the existing tools provide support for real-time editing. Teams must rely on external version control systems and manual coordination to manage shared development, introducing overhead and increasing the risk of integration errors that would be avoided by a collaboration aware environment.

    \item \textbf{Lack of a Centralised System State:} Since tools are local-first, there is no single representation of a system's current state. This makes it difficult to maintain a consistent view of the system across a distributed team to integrate the development environment into automated deployment and monitoring pipelines.

    \item \textbf{Limited Accessibility:} Engineers cannot access a project without access to a fully configured workstation. This is a practical limitation in industrial settings, where on-site diagnosis of a running application may be necessary from a device in which no local tooling has been installed and the project is not set up.

    % TODO remote deployment
\end{itemize}

Taken together, these gaps suggest that the ecosystem is well positioned to benefit from a lightweight, browser-accessible development environment that prioritises collaboration, low setup overhead and integration with modern deployment workflows, without sacrificing the domain-specific features that \textit{IEC 61499} engineers depend on. %TODO maybe dont mention the solution here
