\chapter{Introduction} \label{chap:introduction}

\section{Context}

% Context CPPS and FBD
\ac{CPPS} \cite{monostori_cyber-physical_2014} are complex systems that integrate physical processes with digital technologies to optimise production processes. Several software engineering approaches and technologies are utilised in the development and operation of \ac{CPPS}. One of such approaches is \ac{FBD} \cite{cho_function_1993}, which is a graphical programming language used in industrial automation and control systems. \ac{FBD} uses graphical function blocks to represent the logic and control of a system. The function blocks are connected in a network to form a control program. One of the benefits of \ac{FBD} is that it provides a visual representation of the control logic, which can be easier to understand and debug than traditional textual programming languages. \ac{FBD} is also modular, which means that function blocks can be reused in multiple programs, reducing development time and improving code maintainability. Also, \ac{FBD} is often used in conjunction with \acp{PLC} to control machinery and processes in manufacturing and other industrial applications \cite{pinto_digital_2023,pang_formal_2015}. \ac{FBD} programs can be created and edited using specialised software tools that allow for the creation and manipulation of function blocks \cite{zoitl_developing_2013}.

% Context IEC 61499
% TODO More info about it
In industrial automation, it is common to use the IEC 61499 standard, which defines an event-driven \ac{FBD} way to design control software, especially for distributed systems \cite{sorokin_novel_2022}.

% Context RTE and IDE
% TODO More info about RTE
% TODO Add references
While \acp{RTE} enable the execution of function blocks compliant with the standard, \acp{IDE} are used to design, create, and deploy function block pipelines to \ac{CPPS}. These \acp{IDE} offer features and tools that facilitate the development, testing, and deployment of IEC 61499-compliant control systems. They provide graphical editors for creating function block diagrams, simulation capabilities, and additional functionalities to streamline the development process.

\section{Motivation}

% TODO References
% TODO Write more
% TODO first phrase sucks
The evolution of \ac{CPPS} has increased the demand for flexible and intelligent control solutions capable of integrating physical processes with digital technologies. \ac{FBD} and the IEC 61499 standard offer a structured, modular approach to building such distributed automation systems. To fully leverage these benefits, development tools must evolve alongside industrial needs, supporting more accessible, collaborative, and scalable workflows.

\section{Problem}

% TODO References
% TODO Write more
Existing IEC 61499 development environments are predominantly desktop-based and lack modern capabilities such as remote access, real-time collaboration, and seamless integration with cloud infrastructures. These limitations hinder efficient development, maintenance, and deployment of control applications within distributed \ac{CPPS}. As a result, current tools fall short in supporting the flexibility, scalability required by certain modern industrial automation systems. Their complexity further complicates their usage for more basic needs.

\section{Research Questions}

% TODO Add more sub questions
% TODO Style this better
This research seeks to investigate how modern web technologies can be leveraged to enhance the development, deployment, and operation of IEC 61499-compliant systems. This study is guided by the following main research question:
\begin{quote}
How can web-based \acp{IDE} enhance the development, deployment, and collaboration processes for IEC 61499 applications in distributed \ac{CPPS}, addressing the limitations of traditional desktop-based environments?
\end{quote}
To further refine the scope, the following sub-questions are proposed:
\begin{itemize}
% TODO reference that paper with the full architeture system
  \item What architectural and technological requirements must be considered when designing a web-based \acp{IDE} for IEC 61499?
  % TODO bro a solocao esta na pergunta. a mariana nao aprova
  \item In what ways do web-based \acp{IDE} improve collaboration, accessibility, and scalability compared to existing desktop tools?
\end{itemize}
