\chapter{Proposed Solution} \label{chap:proposed-solution}
To address the limitations identified in existing \textit{IEC 61499} development environments, this dissertation proposes the design and implementation of a lightweight, web-based \ac{IDE} for modelling \textit{IEC 61499}-compliant applications. This work seeks to bridge the gap between modern software engineering practices and the current state of industrial automation tooling for \ac{CPPS}. A web-based \ac{IDE} solves many of the problems mentioned in \Cref{chap:state-of-the-art}, addressing the identified gap.

The main reason to propose using web technologies is to solve the issues that affect any local-first development environment. By having the \ac{IDE} be hosted by a centralised server, the following is achieved:

% TODO the state of the art needs to mention each of these
% TODO a real time collaboration example is excalidraw or other draw programs that use json underneath
\begin{itemize}
  \item \textbf{Ease of setup}. Instead of each team member having to install a resource heavy program and set it up with the team's code and configurations, a single server is set up and the team members simply have to connect to it via their browser.
  \item \textbf{Ease of deployment}. Only the server needs to be aware of all of the target \acp{RTE} and have connection to them. When a team member wants to deploy, the request will be routed through the server, acting as the middleware, removing the need for each development machine to have connection with the target \acp{RTE}.
  \item \textbf{Real-time collaboration}. All data is kept and directly modified in the server itself, instead of requiring a periodic synchronisation system like \textit{git}. \ac{FBD} coded programs, since they are meant for visual editing, need to be stored in formats that don't easily handle merge conflicts. Real-time combats this by keeping track of atomic changes, instead of blocks of text. % TODO ref about git merge conflicts in these types of files
\end{itemize}

\section{Approach}
The approach begins with the derivation of functional and non-functional requirements from the \textit{IEC 61499} standard and an analysis of existing desktop-based tools. Core requirements include support for function block types, event and data connections, and deployment to compliant runtime environments, namely \textit{DINASORE}. Non-functional requirements emphasise platform independence, reduced complexity, usability, and extensibility, addressing the resource intensity and accessibility issues of current solutions. The use of modern web technologies further allows the inclusion of collaborative features and scalable workflows, which are largely absent from current \textit{IEC 61499} tools.

The effectiveness of the approach is assessed through the implementation of a functional prototype and its direct comparison with established desktop-based \acp{IDE} like \textit{4diac}, focusing on usability, accessibility, and workflow efficiency. This evaluation demonstrates the feasibility of applying web-based principles to \textit{IEC 61499} development and their potential benefits for distributed \ac{CPPS}.

% TODO need to explain a lot more of what i want. like a whole page

\section{Methods}
This section describes the research methodology adopted to address the research questions defined in this dissertation. Given the practical and design-oriented nature of the problem, the methodology focuses on the design, implementation, and evaluation of a piece of software. The chosen methods are appropriate for investigating how web-based technologies can enhance \textit{IEC 61499} development environments, as they allow both the realisation of a concrete solution and the systematic assessment of its capabilities relative to existing tools.

\subsection{Research Design}
This research follows a qualitative, design-oriented research approach. The core of the research consists of the design and implementation of a functional prototype of the solution. Rather than aiming for statistical generalisation, the study focuses on demonstrating feasibility and identifying qualitative improvements over existing desktop-based environments. The prototype serves as a research artifact used to explore architectural decisions, usability considerations, and workflow implications.

\subsection{Evaluation}
Evaluation is conducted through qualitative analysis and comparative assessment against established \textit{IEC 61499} \acp{IDE}. Criteria such as accessibility, usability, feature coverage, and support for collaborative workflows are used to assess how effectively the proposed solution addresses the limitations identified in the literature and state of the art.

\section{Workplan}
\Cref{fig:workplan} outlines the planned work for this dissertation. The schedule is organised into three main phases: Experimentation, Code Development, and Deliveries. During the Experimentation phase, there will be a focus on studying the standard and running experiments to establish a solid technical foundation. The Code Development phase overlaps partially with Experimentation, ensuring that the software evolves in parallel with the experimental insights. Finally, the Deliveries phase spans the entire project duration and includes writing the dissertation and preparing the final presentation, ensuring that all outputs are documented and communicated effectively.

\begin{figure}[h]
\begin{ganttchart}[
  x unit=0.6mm,
  y unit chart=0.6cm,
  group height=0.2mm,
  bar height=0.2mm,
  time slot format=isodate
]{2026-02-01}{2026-07-31}
  \gantttitlecalendar{year, month=shortname} \\

  \ganttgroup{Experimentation}{2026-02-01}{2026-04-01} \\
  \ganttbar{Framework Selection}{2026-02-01}{2026-03-01} \\
  \ganttbar{\textit{IEC 61499} study}{2026-02-01}{2026-03-15} \\
  \ganttbar{\ac{RTE} Experimentation}{2026-02-15}{2026-04-01} \\
  \ganttnewline

  \ganttgroup{Code Development}{2026-03-01}{2026-07-15} \\
  \ganttbar{\ac{RTE} Communication}{2026-03-01}{2026-05-01} \\
  \ganttbar{\ac{UI} Development}{2026-04-01}{2026-07-15} \\
  \ganttbar{Advanced Features}{2026-05-15}{2026-07-15} \\
  \ganttnewline

  \ganttgroup{Deliveries}{2026-02-01}{2026-07-31} \\
  \ganttbar{Write Dissertation}{2026-02-01}{2026-07-31} \\
  \ganttbar{Prepare Presentation}{2026-06-01}{2026-07-31} \\
\end{ganttchart}
\caption{Gantt chart timeline for the proposed work}
\label{fig:workplan}
\end{figure}
