\chapter{Conclusion} \label{ch:conclusion}
This dissertation set out to investigate whether modern web technologies and the shift to a middleware driven paradigm could meaningfully improve the development, deployment and operation of \textit{IEC 61499}-compliant distributed automation systems. The answer, backed up by functional, performance, collaboration and usability evaluations, is affirmative.

\section{Contributions and Key Findings}
The central contribution of this work is \textit{PTERO}, a web-based \ac{IDE} for \textit{IEC 61499} function block development. Rather than replicating the local-first architecture that characterises existing tools, \textit{PTERO} adopts a centralised server model in which the browser serves as the sole client-side requirement. This architectural decision has other benefits, such as eliminating per workstation installation and configuration, centralising the deployment state and removing the need for individual workstations to maintain direct network access to target \acp{RTE}.

Beyond the architectural novelty, this dissertation contributes a concrete analysis of the gaps in the existing \textit{IEC 61499} tooling ecosystem and demonstrates, through direct comparison with \textit{4DIAC-IDE}, that those gaps can be addressed without sacrificing functional correctness. Both systems produced identical deployment command sequences and equivalent runtime behaviour in \textit{DINASORE}, confirming that \textit{PTERO}'s abstraction layer preserves full compatibility with the underlying standard and execution infrastructure.

The integration of real-time collaborative editing via a \ac{CRDT} framework represents a further contribution of significance. The collaboration evaluation demonstrated that even resource-constrained hardware, such as the \textit{Raspberry Pi} used throughout testing, can serve as a viable collaboration server for small engineering teams. On more capable hardware, the system scaled reliably to thousands of concurrent clients, validating the architectural scalability of the approach. This stands in contrast to the version control workflows that \textit{4DIAC-IDE} teams must rely upon, which were consistently identified by study participants as a source of friction, merge conflicts and lost productivity.

The usability evaluation reinforced these findings with strong empirical support. The results confirm that the architectural improvements introduced by the web-based model translate directly into a measurably better user experience.

Taken together, the contributions of this dissertation demonstrate that the adoption of web technologies for \textit{IEC 61499} tooling is a design decision that fundamentally restructures the development workflow in ways capable of benefiting accessibility, collaboration and usability.

\section{Impact in the Field}
The findings of this work carry implications that extend beyond \textit{PTERO} itself and offer insights that the wider \textit{IEC 61499} and industrial automation tooling community may draw upon.

The most transferable lesson concerns the treatment of collaboration as an architectural concern rather than a feature to be added after the fact. The difficulties participants reported with \textit{4DIAC-IDE}'s collaboration model arose not from a lack of effort on the part of its developers, but from the inherent limitations of a local-first architecture applied to structured data. Future \acp{IDE} that wish to support collaborative workflows would benefit from adopting \ac{CRDT} based shared state models, synchronising their shared document over a background service rather than through periodic file exchange.

\section{Limitations}
It is important to acknowledge the boundaries of the current prototype. \textit{PTERO} was validated against \textit{DINASORE}. While the management protocol shared between \textit{DINASORE} and \textit{4DIAC FORTE} means that extension to other compliant runtimes is architecturally straightforward, it has not been empirically verified in this work. Similarly, the usability study, while yielding strongly positive results, involved a participant group drawn exclusively from an academic context. Broader validation with practitioners from industrial settings would be necessary before drawing conclusions about the tool's suitability for production deployments.

Furthermore, the current implementation does not yet support the full breadth of \gls{IEC} 61499 constructs which would be required for more complex industrial applications.

\section{Future Work}
Several directions present themselves as natural extensions of this work. The most immediate is support for additional \acp{RTE} beyond \textit{DINASORE}, which would significantly broaden the tool's applicability.

From a collaboration standpoint, the current implementation propagates all graph changes to all connected clients without any notion of user identity or access control. Introducing user authentication, role-based editing permissions and change attribution would make \textit{PTERO} more appropriate for team environments where auditability and access management are required.

On the infrastructure side, the current deployment model assumes that the \textit{PTERO} server has persistent access to all registered \textit{DINASORE} instances. Future work could explore the use of containerised deployment \cite{nikolakis_containerized_2020} and integration with \ac{CI} pipelines, enabling \textit{PTERO} to participate in automated testing and continuous workflows for \textit{IEC 61499} applications.

Finally, a wider usability study conducted in an industrial setting, in which engineering teams adopt \textit{PTERO} over the course of a real development project, would provide much richer empirical evidence than the controlled evaluation performed here.
